Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-18 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Finishing the discussion of results

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use, and
reasonable performance. But hierarchy, this is, the capability of preferentially using resources in the same organization, is a feature that could be either disregarded or included within adaptivity, which leaves us with eight desirable properties.
% (Paloma) Re-write. For example, "But is hierarchy, this is, the capability of preferentially using resources in the same organization, the one that could be either disregarded or included in adaptivity, which leaves us with eight desirable properties."
% (Paloma) And I re-wrote what I proposed yesterday.

@@

These two experiments establish a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing to have more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which this 
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addressess (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: This experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even if that is the case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen, 1 out of roughly 5 or roughly 20, depending on what baseline
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

>>>>

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use, and
reasonable performance. But hierarchy, this is, the capability of
preferentially using resources in the same organization, is a feature
that could be either disregarded or included within adaptivity, which
leaves us with eight desirable properties. 
% (Paloma) Re-write. For example, "But is hierarchy, this is, the
% capability of preferentially using resources in the same
% organization, the one that could be either disregarded or included
% in adaptivity, which leaves us with eight desirable properties." 
% (Paloma) And I re-wrote what I proposed yesterday.
% thanks :-) -JJ

@@

These two experiments establish first a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which this 
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. No special care was taken to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is, in fact negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addressess (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
found previously \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the last two columns, that measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: This experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even if that is the case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen, 1 out of roughly 5 or roughly 20, depending on what baseline
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

This is part due to the fact that the presence of nodes is not
constant, as shown in Figure \cite{fig:ips:rastrigin}, which plots the
number of unique IPs participating in the experiment per minute (left)
and hour (right). The two graphs have the same shape which matches
also the speed in which fitness is decreased in Figure
\ref{fig:puts:rastrigin}. At the beginning there are very few or only
one, but later one there ar peaks of almost 30 unique machine a
minute; these peaks appear again when new announcements are made in
Twitter or any other social networks, to go down to a few remaining
nodes later on. It is interesting to note that, at the most, more than
100 unique nodes were participating in the experiment during one hour;
also that there are sustained peaks during a day of 6-7 computers
every minute. This is due to the fact that the computation is
asynchronous, but also to the fact that different computers stay in
the simulation for a different period of time. This period has been
plotted in Figure \ref{fig:fit:rastrigin}, that plots the number of
{\tt PUT}s per node, ranked by number, along with a fit to the two
distributions we have used before, GEV and Weibull; the parameters of
this fit are shown in Table \ref{tab:puts:ww:f15}. Unlike the previous
experiments, in which GEV fit was relatively good, in this case it is
quite obviously not. However, the fit to the Weibull distribution,
shown at the left of the image, is relatively good. As should be
expected, the shape and scale are different to the one in the previous
experiments with the Trap function, however, the shape is quite
similar (0.40 vs 0.69), being in both cases less than one, which
indicates a decreasing distribution of the same kind.

In general, this experiment confirms what was found in the others in
several aspects: first, there is a regularity in the distribution of
time assigned by volunteers to it, second, there is a good and certain
amount of volunteers that will become part of the distributed
computer, in some cases up to one hundred; also, in some cases there
will be users that will spend several hours, two orders of magnitude
more generations in this case than in the previous one. There is a
clear influence of the time to solution: a short time means that every
experiment will only gather a few volunteers, but also that they will
stay only for a short time. A long time will make users stay for
longer, but if it takes too long, they will eventually lose interest.

We will discuss this and the rest of the findings of the paper in the
conclusions.
